Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-14 conclusions.tex
add new paragraph on sky localization and reorganized intro a bit
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1074 ccdd05d5-56c1-461a-bf98-85f6876523b4

Our future work must also address sky localization accuracy prior to merger as well as
observing strategies.  \citep{Fairhurst2009} derives the area of 90\% localization
confidence in terms of timing uncertainties, but it would be advantageous to use
a galaxy catalog to inform the telescope tiling \citep{galaxy-catalog}.  Because
early detection is likely to be especially effective for nearby sources, the galaxy
catalog technique might be an important ingredient in reducing the fraction of
sky that must be imaged.  Extensive injection campaigns and simulations
incorporating realistic binary merger rates will be necessary in order to fully
understand the prospects for early-warning detection, localization, and \EM\ folluwup
using the techniques we have described.

>>>>

Our future work must also address sky localization accuracy prior to merger as well as
observing strategies.  \citet{Fairhurst2009} derives the area of 90\% localization
confidence in terms of timing uncertainties, but it would be advantageous to use
a galaxy catalog to inform the telescope tiling \citep{galaxy-catalog}.  Because
early detection is likely to be especially effective for nearby sources, the galaxy
catalog technique might be an important ingredient in reducing the fraction of
sky that must be imaged.  Extensive injection campaigns and simulations
incorporating realistic binary merger rates will be necessary in order to fully
understand the prospects for early-warning detection, localization, and \EM\ folluwup
using the techniques we have described.

### 2011-06-14 introduction.tex
add new paragraph on sky localization and reorganized intro a bit
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1074 ccdd05d5-56c1-461a-bf98-85f6876523b4

@@

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection.  Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$1~yr$^{-1}$ will be detectable
within 100~s of merger.  These rates are uncertain and the 5 to 95\% confidence
interval is shown in figure \ref{fig:earlywarning}.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

@@

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm~%
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

\subsection{\EM{} counterparts}

@@

\subsection{Rapid \EM{} followup}

@@

It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$1~yr$^{-1}$ will be detectable
within 100~s of merger. These rates are shown with their substantial uncertainties in figure~\ref{fig:earlywarning}.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

Telescopic followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Unfortunately, for a \GW{} signal with an \SNR{} of
blah, the 90\% confidence area is at least blah deg$^2$ (blah\% of the sky) at
10~s before merger and blah deg$^2$ (blah\% of the sky) at 1~s before merger.
If we can take galactic luminosities to accurately trace mass and thus proxy
for the \CBC{} event rate, it is possible to reduce the localization
uncertainty using a galaxy catalog as a prior probability distribution
\citep{galaxy-catalog}, along with the sky location and luminosity distance
estimate from the \GW{} signal, as was done in S6/VSR3. At Advanced \LIGO{}
\textsc{ns}--\textsc{ns} horizon distances of $\sim$400~Mpc, the galaxy catalog
is fairly isotropic and thus not as useful for downselecting pointings, but
sources that are detectable before merger are likely to be exceptionally
nearby. Thus there will be relatively few galaxies to consider for a given
candidate and the galaxy catalog is likely to be sufficiently complete. This
may allow us to shrink the 90\% confidence area substantially. Even so, it will
be desirable to mobilize a large number of heterogeneous telescopes in order to
guarantee imaging the source at the instant of merger. Because the sky
localization will improve rapidly down to blah deg$^2$ at merger, by the time
that the entire \CBC{} event is over it will be possible to downselect the
images that have already been taken as well as repoint telescopes to more
likely locations.

\subsection{Low-latency \CBC{} search algorithms}

Realizing advance detection of \CBC{}s will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm~%
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

### 2011-06-14 method.tex
add new paragraph on sky localization and reorganized intro a bit
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1074 ccdd05d5-56c1-461a-bf98-85f6876523b4

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in \citet{Marion2004, Buskulic2010}.  We decompose each template
$h_{i}[k]$ into a sum of $S$ non-overlapping templates
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\begin{cases}
		h_i^s[k] & \textrm{if } t^s \leqslant k / f^0 < t^{s+1} \\
		0 & \textrm{otherwise}
	\end{cases}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  The outputs of these new time-sliced filters
form an ensemble of partial \SNR{} streams.  By linearity of the filtering
process, these partial \SNR\ streams can be suitably summed to reproduce the
\SNR\ of the full template.

>>>>

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in \citet{Marion2004} and \citet{Buskulic2010}.  We decompose each template
$h_{i}[k]$ into a sum of $S$ non-overlapping templates
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\begin{cases}
		h_i^s[k] & \textrm{if } t^s \leqslant k / f^0 < t^{s+1} \\
		0 & \textrm{otherwise}
	\end{cases}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  The outputs of these new time-sliced filters
form an ensemble of partial \SNR{} streams.  By linearity of the filtering
process, these partial \SNR\ streams can be suitably summed to reproduce the
\SNR\ of the full template.
